Examining spatial equity and accessibility to a public bicycle share program using a balanced floating catchment area approach

### 2021-03-26 Bike-share-spatial-equity/Bike-share-spatial-equity.Rmd
Revised changes to the manuscript, with reorganizing the introduction and discussion.

Background
===================

Public bicycle share programs (PBSP) have been implemented in over 800 cities worldwide and a great deal has been learned about their typical users [@fishmanBikeshareReviewRecent2016]. In many cities, males use bike share more than females [@breyWantRideMy2017; @nickkarSpatialtemporalGenderLand2019; @reillyGenderDisparitiesNew2020; @wintersWhoAreSuperusers2019] as do younger age cohorts [@breyWantRideMy2017; @buckAreBikeshareUsers2013; @fullerUseNewPublic2011]. However, one study found that bike share users in Washington, DC were more likely to be female [@buckAreBikeshareUsers2013], which suggests that the gender gap among cyclists who use bike share is less disparate than the gap for personal bicycle use [@fishmanBikeshareReviewRecent2016]. There is some evidence that bike share users are less likely to own a car [@buckAreBikeshareUsers2013; @reillyNoncyclistsFrequentCyclists2020]. However, the relationship between income or education and bike share use is less clear. Stations in disadvantaged communities in Chicago generated most of the average annual trips [@qianBikesharingEquityDisadvantaged2020] and individuals from minority or lower socioeconomic status neighbourhoods in Minneapolis-St. Paul used the city's PBSP more [@wangNeighborhoodSociodemographicCharacteristics2019]. Being university educated was also a significant correlate of bike share use in Montreal, Canada [@fullerUseNewPublic2011]. On the other hand, members who reside in minority-concentrated and lower socioeconomic status neighbourhoods use the Minneapolis-St Paul bike share more frequently [@wangNeighborhoodSociodemographicCharacteristics2019]. Financial savings have been found to motivate those on a low income to use bike share [@fishmanBikeshareReviewRecent2016]. Many studies have found that geographic proximity to a bike share station, either living or working near one, is an important determinant of use or having a membership [@fishmanBikeshareReviewRecent2016; @fullerUseNewPublic2011]. This makes sense given that individuals are more likely to use services or programs that they can easily access. Finally, the potential of bike share programs to increase cycling levels has also been explored recently [@hosfordEvaluationImpactPublic2018; @hosfordEvaluatingImpactImplementing2019], but more longitudinal evidence is needed to determine their impact on encouraging more cycling and offering more opportunities for physical activity. On the whole, these findings highlight the need for bike share systems to be more highly accessible for diverse populations in order to increase use beyond the "typical" users, which has been the focus of recent research [@auchinclossDesignBaselineDescription2020; @hullgrassoBikeShareEquity2020; @macarthurAdaptiveBikeShare2020], and to offer more people the option of using sustainable and active transportation.

Several studies have recently explored the equity of PBSP in North American cities by primarily examining who has access to bike share (e.g., differences by demographics or socioeconomic status) and where stations are located. Equity can be achieved in two different ways: equal spatial distribution across a region (e.g., horizontal equity) or greater access for vulnerable or disadvantaged populations (e.g., vertical equity) [@chenExploringEquityPerformance2019]. Both are of interest to researchers and transport planners since they are often linked in that advantage, or conversely disadvantage, has spatial patterns. Using a negative binomial regression model, Qian and Jaller [-@qianBikesharingEquityDisadvantaged2020] estimated ridership in Chicago's PBSP among disadvantaged communities and found some disparities. While annual members in disadvantaged communities have a significantly lower share of trips compared to other areas in the city, they make longer trips. This suggests that they may be using PBSP for work commuting, which points to the importance of ensuring equitable access [@qianBikesharingEquityDisadvantaged2020]. Similar results were found in Philadelphia, where lower income areas generate fewer trips and efforts to increase equity within the program have not been as successful as intended [@caspiBikesharingPhiladelphiaLowerincome2019]. In the case of Seattle, all neighbourhoods had some level of access to dockless bikes but those with higher incomes and more residents of higher education had more bikes [@mooneyFreedomStationSpatial2019]. Babagoli et al. [-@babagoliExploringHealthSpatial2019] found that neighbourhoods in New York City with higher affluence had the greater proportion of Citi Bike stations. Overall, these findings suggest that horizontal equity can be achieved while vertical equity is harder to attain. Exploring transport equity by investigating where bicycle share stations are located, often using neighbourhoods or census tracts as the spatial unit of analysis, can ignore or miss the benefits that may be derived from adjacent zones. Meaning that, stations may be lacking in certain neighbourhoods but there may be stations accessible within a reasonable walking cutoff time. This is where accessibility becomes an important consideration.

@@

Beyond the utility derived from using shared bikes to destinations of value, an important aspect of accessibility associated with PBSPs is the distance an individual must travel to reach a bicycle share station [@kabraBikeShareSystemsAccessibility2020; @wangNewBikeShare2019]. Since the time or distance that members of the population need to reach a bicycle share station decreases the potential of accessing the program and ultimately their use of the program, the location and size (e.g., maximum number of bicycles available) of stations matters. Indeed, distance to bicycle share stations is associated with use [@fullerUseNewPublic2011; @wangNewBikeShare2019] and can be a barrier to using PBSPs [@fishmanBarriersBikesharingAnalysis2014]. Kabra et al. [-@kabraBikeShareSystemsAccessibility2020] found that the majority of bike share usage in Paris comes from areas within 300m of stations. Furthermore, unlike other public amenities with greater ability to adapt to crowding, like health care services, the utility of stations is limited by the number of bicycles that they can hold, which makes crowding effects critical. The program may not necessarily be improved if stations are easy to reach but offer only a small number of bicycles. Likewise, more people may not opt to use the program if the supply of bicycles available at the nearest station is insufficient to meet demand. The location and size of stations is important to increase the utility of this public transportation option for more people, thus achieving vertical equity. Accessibility analyses for PBSPs constitute a positive and evaluation-based approach that also has the potential to inform equity efforts. For instance, Wang and Lindsey [-@wangNewBikeShare2019] investigated whether new or relocated bicycle share stations increased accessibility and use, which offered important insights to improve the performance of the program.

@@

In contrast, floating catchment area (FCA) methods incorporate information on capacity and demand in calculating accessibility. FCA measures have been widely employed in healthcare accessibility research and are typically calculated across two steps. In the first, a ratio of supply to demand at service locations is calculated, such as the number of beds at a hospital divided by the number of people within the catchment area of the hosptital weighted by the distance involved in reaching the facility. Next, these service level ratios are allocated back to the population centres and summarized as a measure of congested accessibility. Thus, this model does a good job of considering potential crowding or compentition for services. While there have been many methodological innovations in FCA methods **(e.g. [Radke & Mu](https://www.tandfonline.com/doi/abs/10.1080/10824000009480538), [Luo & Wang, 2003](https://journals.sagepub.com/doi/10.1068/b29120), [Luo & Qi, 2009](https://www.sciencedirect.com/science/article/pii/S1353829209000574?via%3Dihub), [Wan et al., 2012](https://www.tandfonline.com/doi/full/10.1080/13658816.2011.624987), [Delamater, 2013](https://www.sciencedirect.com/science/article/pii/S1353829213001020?via%3Dihub))**, a recent improvement to this approach was achieved through a simple and intuitive balanced FCA approach that addressed the effects of demand and service inflation from previous FCA approaches [see @paezDemandLevelService2019].

Researchers have also taken different approaches when it comes to aggregation of data, either by using the individual or household as the smallest unit of analysis or larger spatial zones. Previous research on bike share equity has typically used a macro-level approach with aggregate data from entire neighbourhoods or census tracts  [@babagoliExploringHealthSpatial2019; @mooneyFreedomStationSpatial2019; @qianBikesharingEquityDisadvantaged2020; @wangNeighborhoodSociodemographicCharacteristics2019], although there are recent exceptions [@chenExploringEquityPerformance2019; @chenUnobservedHeterogeneityTransportation2021]. This is also true for studies examining correlates of bike share demand [@wangNewBikeShare2019]. Handy and Niemeier [-@handyMeasuringAccessibilityExploration1997] note that using disaggregated data in accessibility analyses provides a more accurate estimate for individuals, which is useful for addressing vertical inequities in PBSP usage.

@@

This paper uses the city of Hamilton, located in Ontario, Canada, as a case study. The city launched a public bicycle share program, Hamilton Bike Share, in March 2015 with 115 stations and 750 bicycles [@hamiltonHamiltonBikeShare2015]. Before June 2020, the program was known as Social Bicycles or SoBi Hamilton. Stations are spaced between 300 and 600 metres apart [@scottRouteChoiceBike2021]. The core service area spans 35 km^2 of the city and roughly 138,000 can reach a bike share hub within 30 minutes of walking [see Figure \ref{fig:hamilton-and-sobi-service-area}]. This represents roughly one fifth of the population in Hamilton according to the 2016 Canadian Census. The program was enthusiastically welcomed in the city - within three weeks of launching, 10,000 trips had been made [@hamiltonHamiltonBikeShare2015]. In 2017, Hamilton Bike Share Inc., the non-profit organization that operates the program, initiated an equity program, the Everyone Rides Initiative (ERI), to remove barriers that may prevent individuals from accessing bike share in Hamilton. An additional 75 bicycles and 15 stations were added to the program which expanded it to more disadvantaged areas in the core service area [see Figure 2]. The program also offers subsidized memberships to individuals who identify as low income. A comparable program can be found in Philadelphia [see @caspiBikesharingPhiladelphiaLowerincome2019]. As of June 2020, the bike share program has 900 bikes, 130 stations [see Figure \ref{fig:sobi-stations-in-hamilton}], and over 26,000 active memberships [@hamiltonHamiltonBikeShare2015].

Hamilton Bike Share has conducted one membership survey in 2018 [@civicplanSoBiHamiltonMembership2017] and the findings from 420 members are broadly in line with the trends that were discussed above [see @fishmanBikeshareReviewRecent2016 for a recent review of the literature]. The majority of respondents live within the core service area and the gender split is typical: 57% of respondents are male and 41% are female. The majority of respondents, both male and female, are between 25 and 34 years of age, but the percentage of male respondents is higher in the subsequent age groups. Respondents use bike share for commuting (40% of trips) or errands and meetings (24% of trips) and nearly 50% of trips have an average length of 11 to 20 minutes. As a result of having a bike share membership, 49% of respondents report that they use their private vehicle less often or much less often but 48% report that their private vehicle use has remained about the same. This suggests that SoBi Hamilton has been useful for certain kinds of trips but not all, meaning that some trips continue to require a private vehicle.

Further research has provided insights about the behaviour and route choice of bike share users in Hamilton. The routes most frequently travelled are longer than the shortest distance route from origin to destination [@luUnderstandingBikeShare2018; @scottRouteChoiceBike2021]. Ridership is influenced by weather conditions, temporal factors such as university terms, employment, and proximity to important destinations like the local university, but is not influenced by population and most transportation infrastructure variables [@scottWhatFactorsInfluence2019]. Bike share users prefer to deter from the shortest distance route to travel on streets with cycling facilities or lower volumes of traffic [@luUnderstandingBikeShare2018] and avoid steep slopes and busy roads [@scottRouteChoiceBike2021]. It is important to note that these studies used daily ridership or GPS data from trips taken before the equity stations were implemented, so these findings are informative with respect to the conventional stations.

Our analysis builds upon a previous and recent study [@hosfordWhoArePublic2018], which found that disadvantaged areas in Hamilton are better served by the city's PBSP compared to other Canadian cities [i.e., Toronto, Vancouver, Montreal, and Ottawa-Gatineau] with PBSPs where advantaged areas have greater access. Hosford and Winters [-@hosfordWhoArePublic2018] acknowledge that "Hamilton stands out in that the lower income neighborhoods are located near the city center and wealthier neighborhoods are in the surrounding suburban areas". Therefore, the core service area for the PBSP in Hamilton is by default in more disadvantaged areas compared to other Canadian cities, but there is also a great deal of variation in income in the city center as well because of the local university and increasing gentrification in the downtown area. In their analysis, Hosford and Winters [-@hosfordWhoArePublic2018] did not differentiate between the conventional stations and those added to increase equity, which we refer to as equity stations in this paper. Therefore, this paper continues their work and makes an important contribution to the growing body of literature that examines equity issues in PBSPs.

@@

Floating catchment area (FCA) methods are an approach commonly used in the healthcare accessibility literature. This approach is more appropriate and informative than calculating the provider-to-population ratio (PPR) which simply divides the level of supply of a service (e.g., the number of bicycle racks at a station) by the population who have access to the service [@paezDemandLevelService2019]. In particular, the Two-Step Floating Catchment Area (2SFCA) method ([Radke & Mu](https://www.tandfonline.com/doi/abs/10.1080/10824000009480538), [Luo & Wang, 2003](https://journals.sagepub.com/doi/10.1068/b29120)) produces flexible catchment areas for populations accessing a service rather than using rigid boundaries like PPR, which provides more useful information because it doesn't assume that people stay within the pre-defined boundaries [@paezDemandLevelService2019]. This is an important property that supports our rationale for applying this method to measure accessibility to Hamilton Bike Share. The City of Hamilton has positioned stations between 300 and 600 metres apart, which would lead to boundaries around these buffers, but it would be reasonable to assume that people are willing to walk beyond this threshold to access other stations if the ones nearest them have no supply of bicycles.

More recently, the _balanced_ floating catchment area approach was developed to address issues with demand and supply inflation that result from the overlapping catchment areas produced by past FCA methods [@paezDemandLevelService2019]. Briefly, the overlapping of catchment areas leads to inflation of population totals and deflation of service levels across a study area and generates inaccurate or misleading accessibility estimates. In contrast, Paez et al. [-@paezDemandLevelService2019] adjust the impedance weights so that both supply and demand are proportionally allocated which eliminates the over-counting of population that leads to distortions in demand and supply. Other benefits of this adjusted method include consideration of competition which can occur when catchment areas overlap, as well as the preservation of system-wide population and level of service. Balanced floating catchment area methods have been used to explore accessibility to health care providers [@paezDemandLevelService2019] and COVID-19 health care services [@pereiraGeographicAccessCOVID192021], but have not yet been used in the cycling literature to explore issues of accessibility.

@@

For this paper, we employ a hybrid place-based and person-based approach to calculating accessibility using disaggregate population data. In their review of accessibility measures, Geurs and van Wee [-@geursAccessibilityEvaluationLanduse2004] highlight the need for greater inclusion of individual spatio-temporal constraints but acknowledge the challenges of acquiring and analyzing person-based data. This comes after Kwan's [-@kwanSpaceTimeIntegral1998] work to show that space-time measures are more capable of capturing interpersonal differences, especially the effect of space-time constraints on individual behaviour, and are more helpful for unraveling gender/ethnic differences. Applying the balanced floating catchment area approach to examine accessibility to Hamilton Bike Share constitutes a location or place-based measure with an individual component by stratifying according to median household income [@geursAccessibilityEvaluationLanduse2004], but conducting a further sensitivity analysis by adjusting distance thresholds would introduce additional spatio-temporal constraints to evaluate equitable accessibility.

@@

[BBBike](https://download.bbbike.org/osm/bbbike/) is an online cycle route planner that interfaces with OpenStreetMap. We extracted OpenStreetMap data for SoBi Hamilton's service area to calculate walking times from each population cell to nearby bike share stations, using a walking distance of 10km and a walking time of 30 minutes as a threshold. A travel time matrix was created with the origins as the coordinates of the population cells and the destinations as the coordinates of the bike share stations within the maximum threshold.

@@

All data for this research were accessed from Open Hamilton^[https://open.hamilton.ca/], an online repository of data curated by the City of Hamilton.

@@

Consensus regarding the distance that individuals are willing to walk to access a PBSP is lacking, therefore the literature on walking behaviour was consulted to determine the thresholds for the sensitivity analysis. Previous studies have found that living within 250 metres [@fullerUseNewPublic2011] and 300 metres [@kabraBikeShareSystemsAccessibility2020] is correlated with bike share use, while other research has found that walking trips are less than 600 metres and rarely more than 1200 metres [@millwardActivetransportWalkingBehavior2013] or a median distance of 650 metres [@larsenQuarterMileReexamining2010]. In Hamilton, Hamilton Bike Share will depict a map at some stations to show the user the locations of the other nearest stations within a five minute walk, which suggests that this is an average distance that people are anticipated to walk. In the present case, we found that congested accessibility calculated using the balanced FCA approach increases with a threshold between two and four minutes, but is then maximized at 5 minutes. Accessibility decreases significantly after eight minutes, which is intuitive given that demand on a limited supply increases as more people can reach each bike share station.

For this reason, we experiment with various walking thresholds by conducting a sensitivity analysis to calculate accessibility at different walking times: 3 minutes, 5 minutes, 10 minutes, and 15 minutes. We categorize these thresholds as minimum, average, maximum, and extreme, respectively. At each threshold, we compare accessibility between the current system and the original system to examine the contribution of the additional equity stations.

@@

### Three Minutes

At the minimum threshold, we find that there are 25.2 bicycle racks per person in the original system.  The addition of equity stations increases this ratio slightly to 25.4 bicycles per person. Figure \ref{fig:figure-6} presents a comparison of accessibility between the systems for the population cells. Accessibility is fairly uniform, with the exception of two small areas near the university and the waterfront park where accessibility is slightly higher.

@@

### Five Minutes

At the average threshold, we find that there are 68.6 bicycle racks per person in the original system.  With the addition of equity stations, there are now 68.8 bicycles per person. Figure \ref{fig:figure-7} presents a comparison of accessibility between the systems. Again, accessibility is fairly uniform, with the exception of one very small area near the waterfront park where accessibility is higher.

@@

### Ten Minutes

>>>>

Introduction
===================

The potential of public bicycle share programs (PBSPs) to increase cycling levels [see @hosfordEvaluationImpactPublic2018; @hosfordEvaluatingImpactImplementing2019] is one of many reasons for implementing such programs in urban areas. As a healthy, inexpensive, and convenient form of public transportation, shared bicycles can encourage individuals to take up cycling for short local trips or first and last mile trips to other public transportation instead of using personal vehicles. Although these programs are available to the general public and ought to be accessible for all individuals who wish to use them, research from PBSPs located in North American and European cities indicates that there are inequities in usage and access that need to be addressed.

## Literature Review

Public bicycle share programs (PBSP) have been implemented in over 800 cities worldwide and a great deal has been learned about their typical users [@fishmanBikeshareReviewRecent2016]. In many cities, males use bike share more than females [@breyWantRideMy2017; @nickkarSpatialtemporalGenderLand2019; @ogilvieInequalitiesUsagePublic2012; @reillyGenderDisparitiesNew2020; @wintersWhoAreSuperusers2019] as do younger age cohorts [@breyWantRideMy2017; @buckAreBikeshareUsers2013; @fullerUseNewPublic2011]. However, one study found that bike share users in Washington, DC were more likely to be female [@buckAreBikeshareUsers2013], which suggests that the gender gap among cyclists who use bike share is less disparate than the gap for personal bicycle use [@fishmanBikeshareReviewRecent2016]. There is some evidence that bike share users are less likely to own a car [@buckAreBikeshareUsers2013; @reillyNoncyclistsFrequentCyclists2020]. However, the relationship between income or education and bike share use is more nuanced. Stations in disadvantaged communities in Chicago were found to generate most of the average annual trips [@qianBikesharingEquityDisadvantaged2020] and individuals from minority or lower socioeconomic status neighbourhoods in Minneapolis-St. Paul used the city's PBSP more [@wangNeighborhoodSociodemographicCharacteristics2019]. Similar findings were reported in London [@ogilvieInequalitiesUsagePublic2012]. Being university educated was a significant correlate of bike share use in Montreal, Canada [@fullerUseNewPublic2011]. Finally, financial savings have been found to motivate those on a low income to use bike share [@fishmanBikeshareReviewRecent2016].

Many studies have found that geographic proximity to a bicycle share station, either living or working near one, is an important determinant of use or having a membership [@fishmanBikeshareReviewRecent2016; @fullerUseNewPublic2011]. This makes sense given that individuals are more likely to use services or programs that they can easily access. Several studies have recently explored the equity of PBSP by primarily examining who has access to bike share (e.g., differences by demographics or socioeconomic status) and where stations are located. It is important to note that equity can be achieved in two different ways: equal spatial distribution across a region (e.g., horizontal equity) or greater access for vulnerable or disadvantaged populations (e.g., vertical equity) [@chenExploringEquityPerformance2019]. Both are of interest to researchers and transport planners since they are often linked in that advantage, or conversely disadvantage, has spatial patterns. Using a negative binomial regression model, Qian and Jaller [-@qianBikesharingEquityDisadvantaged2020] estimated ridership in Chicago's PBSP among disadvantaged communities and found some disparities. While annual members in disadvantaged communities have a significantly lower share of trips compared to other areas in the city, they make longer trips. This suggests that they may be using PBSP for work commuting, which points to the importance of ensuring equitable access [@qianBikesharingEquityDisadvantaged2020]. Similar results were found in Philadelphia, where lower income areas generated fewer trips and it was suggested that efforts to increase equity within the program have not been as successful as intended [@caspiBikesharingPhiladelphiaLowerincome2019]. In the case of Seattle, all neighbourhoods had some level of access to dockless bikes but those with higher incomes and more residents of higher education had more bikes [@mooneyFreedomStationSpatial2019]. Babagoli et al. [-@babagoliExploringHealthSpatial2019] also found that neighbourhoods in New York City with higher affluence had the greatest proportion of Citi Bike stations. Researchers have recommended targeted expansion of stations to areas that use bike share but have inequitable access [@ogilvieInequalitiesUsagePublic2012].

On the whole, existing studies highlight the need for bicycle share systems to be more highly accessible for diverse populations in order to increase use beyond the "typical" users. This has been the focus of recent research [among others, see @auchinclossDesignBaselineDescription2020; @hullgrassoBikeShareEquity2020; @macarthurAdaptiveBikeShare2020]. Offering more people the option of using sustainable and active transportation, particularly those who have lower socioeconomic status and might benefit the most, is a worthy policy goal for cities with PBSPs. Several cities in North America have launched specific programs to address barriers to bike share or to expand service to more deprived areas: London, United Kingdom [@ogilvieInequalitiesUsagePublic2012]; Philadelphia, Pennsylvania [@caspiBikesharingPhiladelphiaLowerincome2019], and Hamilton, Ontario [@hosfordWhoArePublic2018]. Findings to date suggest that horizontal equity can be easier to achieve than vertical equity. However, exploring transportation equity by investigating where bicycle share stations are located, often using neighbourhoods or census tracts as the geographical unit of analysis, can ignore or miss the benefits that may be derived from adjacent zones. Meaning that, stations may be lacking in certain neighbourhoods but there may be stations accessible within a reasonable walking cutoff time. This is where accessibility becomes an important consideration.

@@

Beyond the utility derived from using shared bikes to destinations of value, an important aspect of accessibility associated with PBSPs is the distance an individual must travel to reach a bicycle share station [@kabraBikeShareSystemsAccessibility2020; @wangNewBikeShare2019]. Since the time or distance that members of the population need to reach a bicycle share station decreases the potential of accessing the program and ultimately their use of the program, the location and size (e.g., maximum number of bicycle racks available) of stations matters. Indeed, distance to bicycle share stations is associated with use [@fullerUseNewPublic2011; @wangNewBikeShare2019] and can be a barrier to using PBSPs [@fishmanBarriersBikesharingAnalysis2014]. Kabra et al. [-@kabraBikeShareSystemsAccessibility2020] found that the majority of bike share usage in Paris comes from areas within 300m of stations. Furthermore, unlike other public amenities with greater ability to adapt to crowding, like health care services, the utility of stations is limited by the number of bicycles that they can hold, which makes crowding effects critical. The program may not necessarily be improved if stations are easy to reach but offer only a small number of bicycles. Likewise, more people may not opt to use the program if the supply of bicycles available at the nearest station is insufficient to meet demand. The location and size of stations is important to increase the utility of this public transportation option for more people, thus achieving vertical equity. Accessibility analyses for PBSPs constitute a positive and evaluation-based approach that also has the potential to inform equity efforts. For instance, Wang and Lindsey [-@wangNewBikeShare2019] investigated whether new or relocated bicycle share stations increased accessibility and use, which offered important insights to improve the performance of the program.

@@

In contrast, floating catchment area (FCA) methods incorporate information on capacity and demand in calculating accessibility. FCA measures have been widely employed in healthcare accessibility research and are typically calculated across two steps. In the first, a ratio of supply to demand at service locations is calculated, such as the number of beds at a hospital divided by the number of people within the catchment area of the hospital, weighted by the distance involved in reaching the facility. Next, these service level ratios are allocated back to the population centres and summarized as a measure of congested accessibility. Thus, this model does a good job of considering potential crowding or competition for services. While there have been many methodological innovations in FCA methods [for example, @delamaterSpatialAccessibilitySuboptimally2013; @luoMeasuresSpatialAccessibility2003; @luoEnhancedTwostepFloating2009; @radkeSpatialDecompositionsModeling2000; @wanThreestepFloatingCatchment2012], a recent improvement to this approach was achieved through a simple and intuitive balanced FCA approach that addressed the effects of demand and service inflation from previous FCA approaches [see @paezDemandLevelService2019].

When measuring accessibility, researchers have taken different approaches when it comes to aggregation of data, either by using the individual or household as the smallest unit of analysis or larger spatial zones. Previous research on bike share equity has typically used a macro-level approach with aggregate data from entire neighbourhoods or census tracts  [@babagoliExploringHealthSpatial2019; @mooneyFreedomStationSpatial2019; @qianBikesharingEquityDisadvantaged2020; @wangNeighborhoodSociodemographicCharacteristics2019], although there are recent exceptions [@chenExploringEquityPerformance2019; @chenUnobservedHeterogeneityTransportation2021]. This is also true for studies examining correlates of bike share demand [@wangNewBikeShare2019]. Handy and Niemeier [-@handyMeasuringAccessibilityExploration1997] note that using disaggregated data in accessibility analyses provides a more accurate estimate for individuals, which is useful for addressing vertical inequities in PBSP usage.

@@

## Original System

This paper uses the city of Hamilton, located in Ontario, Canada, as a case study. The city launched a public bicycle share program, Hamilton Bike Share, in March 2015 with 115 stations and 750 bicycles [@hamiltonHamiltonBikeShare2015]. Before June 2020, the program was known as Social Bicycles or SoBi Hamilton but is now called Hamilton Bike Share. Stations are spaced between 300 and 600 metres apart [@scottRouteChoiceBike2021]. The core service area spans 40 square kilometres of the city although it was planned to be 20 square kilometres [@hamiltonsobi2015], and roughly 138,000 can reach a bike share hub within 30 minutes of walking [see Figure \ref{fig:hamilton-and-sobi-service-area}]. This represents roughly one fifth of the population in Hamilton according to the 2016 Canadian Census. The City of Hamilton undertook a large public engagement campaign to validate the locations of bicycle share stations that had been selected and to crowdsource potential locations for additional locations [@hamiltonsobi2014]. Most of the suggested locations were in the east end of the core service area that lack stations or in neighbourhoods not serviced by the PBSP. Inadequate coverage given the size of the Hamilton Bike Share's service area was identified as a problem early on, with transportation planners noting that the small size (i.e., supply of bicycle racks) and low quantity of stations would lead to challenges in balancing the system [@hamiltonsobi2015]. Based on public feedback, 13 stations were added to the original system. The program was enthusiastically welcomed in the city in 2015 - within three weeks of launching, 10,000 trips had been made [@hamiltonHamiltonBikeShare2015].

## Equity Initiative

In 2017, Hamilton Bike Share Inc., the non-profit organization that operates the program, initiated an equity program, Everyone Rides Initiative (ERI), to remove barriers that may prevent individuals from accessing bike share in Hamilton. Additional bicycles and stations were added to the program which expanded it to more disadvantaged areas in the core service area [see Figure 2]. The program also offers subsidized memberships to individuals who identify as low income as well as cycle skills education. A comparable program can be found in Philadelphia [see @caspiBikesharingPhiladelphiaLowerincome2019].

## Current System

As of June 2020, Hamilton Bike Share has 900 bikes, 130 stations [see Figure \ref{fig:sobi-stations-in-hamilton}], and over 26,000 active memberships [@hamiltonHamiltonBikeShare2015]. The core service area remains 40 square kilometres. The PBSP has 13 stations that were added as part of the ERI initiative; we refer to these as "equity stations" throughout the paper while all other stations are "conventional stations". In total, 117 stations are "conventional" and 13 are "equity".

## Membership

Hamilton Bike Share conducted one membership survey in 2018 [@civicplanSoBiHamiltonMembership2017] and the findings from 420 members are broadly in line with the trends that were discussed above [see @fishmanBikeshareReviewRecent2016 for a recent review of the literature]. The majority of respondents live within the core service area and the gender split is expected: 57% of respondents are male and 41% are female. The majority of respondents, both male and female, are between 25 and 34 years of age, but the percentage of male respondents is higher in the subsequent age groups. Respondents use bike share for commuting (40% of trips) or errands and meetings (24% of trips) and nearly 50% of trips have an average length of 11 to 20 minutes. As a result of having a bike share membership, 49% of respondents report that they use their private vehicle less often or much less often but 48% report that their private vehicle use has remained about the same. This suggests that SoBi Hamilton has been useful for certain kinds of trips but not all, meaning that some trips continue to require a private vehicle.

## Relevant Research

Our analysis builds upon a previous and recent study [@hosfordWhoArePublic2018], which found that areas in Hamilton with less advantage are better served by the city's PBSP compared to other Canadian cities [i.e., Toronto, Vancouver, Montreal, and Ottawa-Gatineau] where areas that are less deprived have greater access. Hosford and Winters [-@hosfordWhoArePublic2018] acknowledge that "Hamilton stands out in that the lower income neighborhoods are located near the city center and wealthier neighborhoods are in the surrounding suburban areas". Therefore, the core service area for the PBSP in Hamilton is by default in more disadvantaged areas compared to other parts of the city. However, there is also a great deal of variation in income in the city center because of the local university and increasing gentrification. Hosford and Winters [-@hosfordWhoArePublic2018] took a macro-level and system-wide approach in their analysis by using dissemination areas _across_ the city as the unit of analysis. They did not focus specifically on the core service area and did not differentiate between conventional and equity stations, which is worth exploring further. Therefore, this paper continues their analysis and makes an important contribution to the growing body of literature that examines equity issues in PBSPs.

@@

Floating catchment area (FCA) methods are an approach commonly used in the healthcare accessibility literature. This approach is more appropriate and informative than calculating the provider-to-population ratio (PPR) which simply divides the level of supply of a service (e.g., the number of bicycle racks at a station) by the population who have access to the service [@paezDemandLevelService2019]. In particular, the Two-Step Floating Catchment Area (2SFCA) method [@luoMeasuresSpatialAccessibility2003; @radkeSpatialDecompositionsModeling2000] produces flexible catchment areas for populations accessing a service rather than using rigid boundaries like PPR, which provides more useful information because it doesn't assume that people stay within the pre-defined boundaries [@paezDemandLevelService2019]. This is an important property that supports our rationale for applying this method to measure accessibility to Hamilton Bike Share. The City of Hamilton has positioned stations between 300 and 600 metres apart, but anticipates that hubs will service those living within a 250 metre buffer from the station [@hamiltonsobi2015]. The latter constitutes a normative statement: people ought to be able to access a station in less than 600 metres if they live in the core service area, with usage coming from 250 metres around. However, it is not known how far people are actually willing to travel to reach a station. It would be reasonable to assume that people are willing to walk beyond this threshold to access other stations if the ones nearest them have no supply of bicycles.

More recently, the _balanced_ floating catchment area (FCA) approach was developed to address issues with demand and supply inflation that result from the overlapping catchment areas produced by past FCA methods [@paezDemandLevelService2019]. Briefly, the overlapping of catchment areas leads to inflation of population totals and deflation of service levels across a study area and generates inaccurate or misleading accessibility estimates. In contrast, Paez et al. [-@paezDemandLevelService2019] adjust the impedance weights so that both supply and demand are proportionally allocated which eliminates the over-counting of population that leads to distortions in demand and supply. Other benefits of this adjusted method include consideration of competition which can occur when catchment areas overlap, as well as the preservation of system-wide population and level of service. Balanced floating catchment area methods have been used to explore accessibility to health care providers [@paezDemandLevelService2019] and COVID-19 health care services [@pereiraGeographicAccessCOVID192021], but have not yet been used in the cycling literature to explore issues of accessibility.

@@

For this paper, we employ a hybrid location-based and person-based approach to calculating accessibility using disaggregate population data. In their review of accessibility measures, Geurs and van Wee [-@geursAccessibilityEvaluationLanduse2004] highlight the need for greater inclusion of individual spatio-temporal constraints but acknowledge the challenges of acquiring and analyzing person-based data. This comes after Kwan's [-@kwanSpaceTimeIntegral1998] work to show that space-time measures are more capable of capturing interpersonal differences, especially the effect of space-time constraints on individual behaviour, and are more helpful for unraveling gender/ethnic differences. Applying the balanced floating catchment area approach to allows us to examine accessibility by stratifying according to median household income, which would constitute the individual component of the accessibility measure [@geursAccessibilityEvaluationLanduse2004]. However, conducting a further sensitivity analysis to measure accessibility at different distance thresholds would help to consider potential spatio-temporal constraints. Different people may be willing to travel different distances to access a public bicycle share station.

@@

[BBBike](https://download.bbbike.org/osm/bbbike/) is an online cycle route planner that interfaces with OpenStreetMap. We extracted OpenStreetMap data for SoBi Hamilton's service area to calculate walking times from each population cell to nearby bike share stations, using a walking distance of 10km and a walking time of 30 minutes as a threshold. A travel time matrix was created with the origins as the coordinates of the population cells and the destinations as the coordinates of the bike share stations within the maximum threshold. This provides a more realistic measure of the temporal component by taking infrastructure into account of travel times, rather than using the Euclidean distance from population cell to station.

@@

All data for this research were accessed from Open Hamilton^[https://open.hamilton.ca/], an online and public repository of data curated by the City of Hamilton.

@@

Consensus regarding the distance that individuals are willing to walk to access a PBSP is lacking, but the literature on walking behaviour was consulted to determine the thresholds for the sensitivity analysis. Previous studies have found that living within 250 metres [@fullerUseNewPublic2011] and 300 metres [@kabraBikeShareSystemsAccessibility2020] is correlated with bike share use, while other research has found that walking trips are less than 600 metres and rarely more than 1200 metres [@millwardActivetransportWalkingBehavior2013] or a median distance of 650 metres [@larsenQuarterMileReexamining2010]. In Hamilton, Hamilton Bike Share will depict a map at some stations to show the user the locations of the other nearest stations within a five minute walk, which suggests that this is an average distance that people are expected to walk.

In the present case, we found that congested accessibility calculated using the balanced FCA approach increases with a threshold between two and four minutes, but is then maximized at 5 minutes. Accessibility decreases significantly after eight minutes, which is intuitive given that demand on a limited supply increases as more people can reach each bike share station.

For this reason, we experiment with various walking thresholds by conducting a sensitivity analysis to calculate accessibility at different walking times from population cell to bicycle share station: 3 minutes, 5 minutes, 10 minutes, and 15 minutes. We categorize these thresholds as minimum, average, maximum, and extreme, respectively. At each threshold, we compare accessibility between the current system and the original system to examine the contribution of the additional equity stations.

@@

### Minimum Threshold

With a walking distance of three minutes, we find that there are 25.2 bicycle racks per person in the original system. The addition of equity stations increases this ratio slightly to 25.4 bicycles per person. Figure \ref{fig:figure-6} presents a comparison of accessibility between the systems for the population cells. Accessibility is fairly uniform overall, with the exception of two small areas where accessibility is slightly higher.

@@

### Average Threshold

With a walking distance of five minutes, we find that there are 68.6 bicycle racks per person in the original system. With the addition of equity stations, there are now 68.8 bicycles per person. Figure \ref{fig:figure-7} presents a comparison of accessibility between the systems. Again, accessibility is fairly uniform, with the exception of one very small area. At this threshold, there are more bicycle racks per person than at the minimum threshold. This occurs because fewer stations can be reached when the travel time is 3 minutes or less.

@@

### Maximum Threshold
